Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-03-20 _tex/paper2.tex
Built site for gh-pages

\begin{abstract}
This study revisits Thurstone's law of comparative judgments (CJ) by
addressing two key limitations in traditional approaches. Firstly, it
addresses the overreliance on the assumptions of Thurstone's Case V in
the statistical analysis of CJ data. Secondly, it addresses the apparent
disconnect between CJ's approach to trait measurement and hypothesis
testing. We put forward a systematic approach based on causal analysis
and Bayesian statistical methods, which results in a model that
facilitates a more comprehensive understanding of the factors
influencing CJ experiments while offering a robust statistical
translation. The new model accommodates unequal dispersions and
correlations between stimuli, enhancing the reliability and validity of
CJ's trait estimation, thereby ensuring the accurate measurement and
interpretation of comparative data. The paper highlights the relevance
of this updated framework for modern empirical research, particularly in
education and social sciences. This contribution advances current
research methodologies, providing a robust foundation for future
applications in diverse fields.
\end{abstract}

@@

Numerous studies have documented the effectiveness of CJ in assessing
traits and competencies over the past decade. These studies have
highlighted three aspects of the method's effectiveness: its
reliability, validity, and practical applicability. Research on
reliability suggests that CJ requires a relatively modest number of
pairwise comparisons \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}
to generate trait scores that are as precise and consistent as those
generated by other assessment methods
\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}. In
addition, the evidence suggests that the reliability and time efficiency
of CJ are comparable, if not superior, to those of other assessment
methods when employing adaptive comparison algorithms
\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}.
Meanwhile, research on the validity of CJ scores indicates their
capacity to represent accurately the traits under measurement
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
Moreover, research on its practical applicability highlights CJ's
versatility across both educational and non-educational contexts
\citep{Kimbell_2012, Jones_et_al_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues that can undermine the reliability and validity of CJ's
trait estimates. The first issue arises from the excessive reliance on
Thurstone's Case V assumptions in the statistical analysis of CJ data.
The second stems from the apparent disconnect between CJ's approach to
trait measurement and hypothesis testing. This study addresses these
issues by extending Thurstone's general form through a systematic and
integrated approach based on Causal and Bayesian inference methods.

As a result, the study divides its content into six main sections.
Section~\ref{sec-thurstone_theory} provides an overview of Thurstone's
theory. Section~\ref{sec-theory-issues} discusses the identified issues
in detail. Section~\ref{sec-theoretical} extends Thurstone's general
form to address these challenges. The extension integrates core
theoretical principles alongside key assessment design features relevant
to CJ experiments, such as the selection of judges, stimuli, and
comparisons. Section~\ref{sec-statistical} translates these theoretical
and practical elements into a probabilistic statistical model for the
analysis of dichotomous pairwise comparison data.
Section~\ref{sec-discussion} discusses the findings, limitations,
challenges and explores avenues for future research. Finally,
Section~\ref{sec-conclusion} summarizes the study's key insights.

@@

However, since the individual discriminal processes of the stimuli are
not directly observable, the theory introduces the \emph{law of
comparative judgment}. This law posits that in pairwise comparisons, a
judge perceives the stimulus with a discriminal process positioned
further along the trait continuum as possessing more of the trait
\citep[pp.~251]{Bramley_2008}. This suggests that the relative distance
between stimuli, rather than their absolute positions on the continuum,
likely defines the outcome of pairwise comparisons. Indeed, the theory
assumes that the difference between the underlying discriminal processes
of the stimuli, referred to as the \emph{discriminal difference},
determines the observed dichotomous outcome. Furthermore, the theory
assumes that because the individual discriminal processes form a Normal
distribution on the continuum, the discriminal difference will also
conform to a Normal distribution \citep[pp.~452]{Andrich_1978}. In this
distribution, the mode (mean) represents the relative separation between
the stimuli, and its dispersion indicates the variability of that
separation.

Figure~\ref{fig-discriminal_difference} illustrates the distribution of
the discriminal difference for the hypothetical texts depicted in
Figure~\ref{fig-discriminal_process}. The figure indicates that the
judge perceives Text B as having significantly higher quality than Text
A. This conclusion is supported by two key observations: the positive
difference between their modal discriminal processes
\((T_{B} - T_{A} > 0)\) and the probability area where the discriminal
difference distinctly favors Text B over Text A, represented by the
shaded gray area denoted as \(P(B > A)\). As a result, the dichotomous
outcome of this comparison is more likely to favor Text B over Text A.

\section{Two prominent issues in CJ practice}\label{sec-theory-issues}

Thurstone noted from the outset that his general form, described in
Section~\ref{sec-thurstone_theory}, led to a \emph{trait scaling
problem}. Specifically, the model required estimating more ``unknown''
parameters than the number of available pairwise comparisons
\citep[pp.~267]{Thurstone_1927b}. For instance, in a CJ experiment with
five texts, the general form would require estimating \(20\) parameters:
five modal discriminal processes, five discriminal dispersions, and
\(10\) correlations --one per comparison (see
Table~\ref{tbl-thurstone_cases}). However, a single judge could provide
only \({5 \choose 2} = 10\) unique comparisons, an insufficient data set
to estimate the required parameters.

@@

It is evident that although Thurstone developed Case V to provide a
``rather coarse scaling'' of traits \citep[pp.~269]{Thurstone_1927b} and
cautioned that its use ``should not be made without experimental test''
\citep[pp.~270]{Thurstone_1927b}, the CJ literature has nevertheless
relied the most on it in practice. This preference is particularly
notable given that Case V incorporates the most extensive set of
simplifying assumptions among all the cases
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}}
(see Table~\ref{tbl-thurstone_cases}).

Furthermore, since Thurstone's primary goal was to define a
psychological scale and ``allocate the compared stimuli on this
continuum'' \citep[pp.~269]{Thurstone_1927b}, he did not provide
guidance on how to use these measurements when, for instance,
researchers aim to test hypotheses related to the measured traits.
Although the CJ practice has circumvented this issue by using scores'
point estimates or their transformations for such analyses, a
fundamental question remains: Is this method appropriate for addressing
these inquiries?

Thus, this section discusses these two prominent issues, as they could
undermine the reliability and validity of CJ's trait estimates
\citep[pp.~2]{Perron_et_al_2015}. Specifically,
Section~\ref{sec-theory-issue1} examines the the excessive reliance on
Thurstone's Case V assumptions in the statistical analysis of CJ data.
Moreover, Section~\ref{sec-theory-issue2} focuses on the apparent
disconnect between the approaches to trait measurement and hypothesis
testing in CJ.

\subsection{The overreliance on Case V in the statistical analysis of CJ
data}\label{sec-theory-issue1}

As previously discussed, despite relying on the most extensive set of
simplifying assumptions
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}},
Case V remains the most widely used case in the CJ literature. This
popularity stems mainly from its simplified statistical representation
in the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959}. The BTL model mirrors the
assumptions of Case V, with one notable distinction: whereas Case V
assumes a Normal distribution for the stimuli discriminal processes, the
BTL model uses the more mathematically tractable Logistic distribution
\citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). As noted by
\citet[pp.~16]{vanderLinden_et_al_2017_I}, this substitution has little
impact on the model's estimation or interpretation because the
discriminal process scale is arbitrary up to a non-monotonic
transformation. Furthermore, this minimal impact is reinforced by the
fact that the Normal and Logistic distributions exhibit analogous
statistical properties, differing only by a scaling factor of
approximately \(1.7\).

However, as mentioned earlier, Thurstone originally developed Case V to
provide a ``rather coarse scaling'' of traits
\citep[pp.~269]{Thurstone_1927b}, prioritizing statistical simplicity
over precision in trait measurement \citep[pp.~677]{Kelly_et_al_2022}.
He explicitly warned against its untested application, stating that its
use ``should not be made without (an) experimental test''
\citep[pp.~270]{Thurstone_1927b}. Furthermore, he acknowledged that some
assumptions could prove problematic when researchers assess complex
traits or heterogeneous stimuli \citep[pp.~376]{Thurstone_1927a}.
Consequently, given that modern CJ applications frequently involve such
traits and stimuli, two main assumptions of Case V and, by extension, of
the BTL model may not consistently hold in theory or practice, namely
the assumption of equal dispersion and zero correlation between stimuli.

@@

Thurstone argued that the assumption of equal dispersions may not apply
when researchers assess complex traits or heterogeneous stimuli
\citep[pp.~376]{Thurstone_1927a}, as these traits and stimuli can
introduce judgment discrepancies due to their unique characteristics
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Indeed, the CJ literature may already provide evidence of this violation
through ``misfit'' statistics. For example, labeling texts as
\emph{misfits} indicates that comparisons involving these texts result
in more judgment discrepancies than those involving other texts
\citetext{\citealp[pp.~11]{Pollitt_2004}; \citealp[pp.~290]{Pollitt_2012b}}.
These discrepancies imply that the discriminal differences for
\emph{misfit} texts have broader distributions, further suggesting that
their discriminal processes exhibit more dispersion than those of other
texts. This idea can be illustrated using
Figure~\ref{fig-discriminal_process}, where text B shows a wider
discriminal dispersion than text A. As a result, the texts' discriminal
processes overlap more, making it difficult for a judge to determine
which text possesses more of the trait.

A similar line of reasoning applies to the concept of ``misfit'' judges,
whose evaluations deviate substantially from the shared consensus due to
the unique characteristics of the stimuli or the judges themselves
\citetext{\citealp[pp.~164-165]{Pollitt_2012a}; \citealp[pp.~289-290]{Pollitt_2012b}; \citealp[pp.~4]{vanDaal_et_al_2016}; \citealp[pp.~20]{Goossens_et_al_2018}}.
These ``misfit'' judges and their associated deviations can give rise to
additional statistical and measurement issues, which we discuss in more
detail in Section~\ref{sec-theory-issue1b}.

According to the theory, discrepancies in the discriminal dispersions of
stimuli shape the distribution of the discriminal difference, exerting a
direct influence on the outcome of pairwise comparisons. A thought
experiment can illustrate this idea. In this experiment, researchers can
observe the discriminal processes for the texts depicted in
Figure~\ref{fig-discriminal_process}. Furthermore, the experiment
assumes that the discriminal dispersion for Text A remains constant and
that the texts are uncorrelated \((\rho=0)\).
Figure~\ref{fig-dispersion} reveals that an increase in the uncertainty
associated with the perception of Text B compared to Text A
\((\sigma_{B} - \sigma_{A})\), broadens the distribution of their
discriminal difference. This broadening affects the probability area
where the discriminal difference distinctly favors Text B over Text A,
expressed as \(P(B>A)\), ultimately influencing the comparison outcome.
Additionally, the figure reveals that when the discriminal dispersions
of the texts are equal \((\sigma_{B} - \sigma_{A}=0)\), the discriminal
difference is more narrow compared to situations where their dispersions
differ. Consequently, the discriminal difference is more likely to favor
Text B over Text A (shaded gray area).

@@

In experimental practice, however, the thought experiment occurs in
reverse. Researchers first observe the comparison outcome and then use
the BTL model to infer the discriminal difference between stimuli and
their respective discriminal processes \citep[pp.~373]{Thurstone_1927a}.
Consequently, the ability of the outcome to reflect \emph{true}
differences between stimuli largely depends on the validity of the
model's assumptions \citep[pp.~150]{Kohler_et_al_2019}, particularly the
assumption of equal dispersions. For example, when researchers correctly
assume equal dispersions between texts, the BTL model estimates a
discriminal difference distribution that accurately represents the
\emph{true} discriminal difference between them. This scenario is
illustrated in Figure~\ref{fig-dispersion}, when the model's discriminal
difference distribution aligns with the \emph{true} discriminal
difference distribution, represented by the thick continuous line
corresponding to \(\sigma_{B}-\sigma_{A}=0\). The accuracy of this
discriminal difference then ensures reliable estimates for the texts'
discriminal processes.

However, erroneously assuming equal dispersions between stimuli can give
rise to significant statistical and measurement issues. Specifically,
the model may overestimate the trait's reliability, that is, the degree
to which the outcome accurately reflects the \emph{true} discriminal
differences between stimuli. This overestimation, in turn, results in
spurious conclusions about these differences
\citep[pp.~370]{McElreath_2020} and, by extension, about the underlying
discriminal processes of stimuli. Figure~\ref{fig-dispersion} also
illustrates this scenario when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the \emph{true} discriminal
difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\). Additionally, if researchers recognize
that misfit statistics highlight these critical differences in
dispersions, the conventional CJ practice of excluding stimuli based on
these statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018}
can unintentionally discard valuable information, introducing bias into
trait estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}. The
direction and magnitude of these biases are often unpredictable, as they
depend on which stimuli are excluded from the analysis.

@@

In contrast, Thurstone assumed that stimuli were uncorrelated because
judges' biases, arising from two opposing and equally weighted effects
occurring during the pairwise comparisons, canceled each other out
\citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978} provided a
mathematical demonstration of this cancellation using the BTL model
under the assumption of discriminal processes with additive biases.
However, the assumption of zero correlation is not met in practice in at
least two scenarios. First, when intricate aspects of multidimensional,
complex traits or heterogeneous stimuli influence judges' perceptions,
generating biases that resist cancellation. Second, when an additional
hierarchical structure is relevant to the stimuli.

In the first scenario, research on text quality suggests that when
judges evaluate multidimensional, complex traits or heterogeneous
stimuli, they often rely on various intricate characteristics of the
stimuli to form their judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
additional relevant characteristics are unlikely to be equally weighted
or opposing. As a result, they may exert an uneven influence on judges'
perceptions, creating biases in their judgments that resist
cancellation. For example, this could occur when a judge assessing the
argumentative quality of a text places more weight on its grammatical
accuracy than other judges, thereby favoring texts with fewer errors but
weaker arguments. Furthermore, since the discriminal difference of the
stimuli becomes an observable outcome only through the judges'
perceptions, these biases could ultimately introduce dependencies
between the stimuli \citep[pp.~346]{vanderLinden_et_al_2017_II}. While
direct evidence for this particular scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
supporting the notion that the factors influencing pairwise comparisons
may not always cancel out.

In the second scenario, the shared context or inherent connections
introduced by additional hierarchical structures may create dependencies
between stimuli, a statistical phenomenon known as clustering
\citep{Everitt_et_al_2010}. Although the CJ literature acknowledges the
existence of such hierarchical structures, the statistical approaches to
account for this additional source of dependence have been insufficient.
For instance, when CJ data incorporates multiple samples of stimuli from
the same individuals, researchers frequently rely on (averaged) point
estimates of the BTL scores to conduct subsequent analyses and tests at
the individual hierarchical level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
However, this approach can introduce additional statistical and
measurement issues, which we discuss in greater detail in
Section~\ref{sec-theory-issue2}.

According to the theory, the correlation \(\rho\) measures how much the
judges' perception of a specific trait in one stimulus depends on their
perception of the same trait in another. As with the discriminal
dispersions, this correlation shapes the distribution of the discriminal
difference, directly impacting the outcomes of pairwise comparisons. A
similar thought experiment, as the one depicted in
Section~\ref{sec-theory-issue1a}, can illustrate this idea. The
experiment now assumes that the discriminal dispersions for both texts
remain constant. Figure~\ref{fig-correlation} reveals that as the
correlation between the texts increases, the distribution of their
discriminal difference becomes narrower. This narrowing affects the area
under the curve where the discriminal difference distinctly favors Text
B over Text A, denoted as \(P(B > A)\), thus influencing the comparison
outcome. Furthermore, the figure shows that when two texts are
independent or uncorrelated \((\rho=0)\), their discriminal difference
is less narrow compared to scenarios where the texts are highly
correlated. Consequently, the discriminal difference is less likely to
favor Text B over Text A (shaded gray area).

>>>>

\begin{abstract}
This study revisits Thurstone's law of comparative judgment (CJ),
focusing on two prominent issues of traditional approaches. First, it
critiques the heavy reliance on Thurstone's Case V assumptions and, by
extension, the Bradley-Terry-Luce (BTL) model when analyzing CJ data.
Specifically, the study raises concerns about the assumptions of equal
discriminal dispersions and zero correlation between the stimuli. While
these assumptions simplify the trait measurement process, they may fail
to capture the complexity of CJ data, potentially leading to unreliable
and inaccurate trait estimates. Second, the study highlights the
apparent disconnect between CJ's trait measurement and hypothesis
testing processes. Although separating these processes simplifies the
analysis of CJ data, it may also undermine the reliability of various
statistical results derived from the measurement.

To address these issues, the study extends Thurstone's general form
using a systematic and integrated approach based on Causal and Bayesian
inference methods. This extension integrates core theoretical principles
alongside key assessment design features relevant to CJ experiments,
such as the selection of judges, stimuli, and comparisons. It then
translates these elements into a probabilistic statistical model for
analyzing dichotomous CJ data, overcoming the rigid assumptions of Case
V and the BTL model.

Finally, the study emphasizes the relevance of this extension for
contemporary empirical CJ research, particularly stressing the need for
bespoke CJ models tailored to the specific assumptions of the
experiments and data. It also lays the foundation for broader
applications, encouraging researchers across the social sciences to
adopt more robust and interpretable methodologies.
\end{abstract}

@@

Numerous studies have documented the effectiveness of CJ in assessing
traits and competencies over the past decade. These studies have
highlighted three aspects of the method's effectiveness: its
reliability, validity, and practical applicability. Research on
reliability suggests that CJ requires a relatively modest number of
pairwise comparisons \citep{Verhavert_et_al_2019, Crompvoets_et_al_2022}
to generate trait scores that are as precise and consistent as those
generated by other assessment methods
\citep{Coertjens_et_al_2017, Goossens_et_al_2018, Bouwer_et_al_2023}. In
addition, the evidence suggests that the reliability and time efficiency
of CJ are comparable, if not superior, to those of other assessment
methods when employing adaptive comparison algorithms
\citep{Pollitt_2012b, Verhavert_et_al_2022, Mikhailiuk_et_al_2021}.
Meanwhile, research on validity indicates the capacity of CJ scores to
represent accurately the traits under measurement
\citep{Whitehouse_2012, vanDaal_et_al_2016, Lesterhuis_2018_thesis, Bartholomew_et_al_2018, Bouwer_et_al_2023}.
Lastly, research on its practical applicability highlights CJ's
versatility across both educational and non-educational contexts
\citep{Kimbell_2012, Jones_et_al_2015, Bartholomew_et_al_2018, Jones_et_al_2019, Marshall_et_al_2020, Bartholomew_et_al_2020, Boonen_et_al_2020}.

Nevertheless, despite the increasing number of CJ studies, research in
this domain remains unsystematic and fragmented, leaving several
critical issues unresolved. This study identifies and discusses two
prominent issues of traditional approaches that can undermine the
reliability and validity of CJ's trait estimates
\citep[pp.~2]{Perron_et_al_2015}. First, it critiques the heavy reliance
on Thurstone's Case V assumptions \citep{Thurstone_1927b} and, by
extension, the Bradley-Terry-Luce (BTL) model
\citep{Bradley_et_al_1952, Luce_1959} when analyzing CJ data.
Specifically, the study raises concerns about the assumptions of equal
discriminal dispersions and zero correlation between the stimuli. While
these assumptions simplify statistical modeling, they may fail to
capture the complexity of CJ data, potentially leading to unreliable and
inaccurate trait estimates. Second, the study highlights the apparent
disconnect between CJ's trait measurement and hypothesis testing
processes. Although separating these processes simplifies the analysis
of CJ data, it may also undermine the reliability of various statistical
results derived from the measurement.

To address these issues, the study extends Thurstone's general form
using a systematic and integrated approach based on Causal and Bayesian
inference methods. As a result, the study divides its content into six
main sections. Section~\ref{sec-thurstone_theory} provides an overview
of Thurstone's theory. Section~\ref{sec-theory-issues} discusses the
identified issues in detail. Section~\ref{sec-theoretical} extends
Thurstone's general form to address these challenges. The extension
integrates core theoretical principles alongside key assessment design
features relevant to CJ experiments, such as the selection of judges,
stimuli, and comparisons. Section~\ref{sec-statistical} translates these
theoretical and practical elements into a probabilistic statistical
model to analyze dichotomous pairwise comparison data.
Section~\ref{sec-discussion} discusses the findings, limitations, and
challenges and explores avenues for future research. Finally,
Section~\ref{sec-conclusion} summarizes the study's conclusions.

@@

However, since the individual discriminal processes of the stimuli are
not directly observable, the theory introduces the \emph{law of
comparative judgment}. This law posits that in pairwise comparisons, a
judge perceives the stimulus with a discriminal process positioned
further along the trait continuum as possessing more of the trait
\citep[pp.~251]{Bramley_2008}. This suggests that pairwise comparison
outcomes depend on the relative distance between stimuli, not their
absolute positions on the continuum. Indeed, the theory assumes that the
difference between the underlying discriminal processes of the stimuli,
referred to as the \emph{discriminal difference}, determines the
observed dichotomous outcome. Furthermore, the theory assumes that
because the individual discriminal processes form a Normal distribution
on the continuum, the discriminal difference will also conform to a
Normal distribution \citep[pp.~452]{Andrich_1978}. In this distribution,
the mode (mean) represents the relative separation between the stimuli,
and its dispersion indicates the variability of that separation.

Figure~\ref{fig-discriminal_difference} illustrates the distribution of
the discriminal difference for the hypothetical texts depicted in
Figure~\ref{fig-discriminal_process}. The figure indicates that the
judge perceives Text B as having significantly higher quality than Text
A. Two key observations support this conclusion: the positive difference
between their modal discriminal processes \((T_{B} - T_{A} > 0)\) and
the probability area where the discriminal difference distinctly favors
Text B over Text A, represented by the shaded gray area denoted as
\(P(B > A)\). As a result, the dichotomous outcome of this comparison is
more likely to favor Text B over Text A.

\section{Two Prominent Issues in Traditional CJ
Practice}\label{sec-theory-issues}

Thurstone noted from the outset that his general form, described in
Section~\ref{sec-thurstone_theory}, led to a \emph{trait scaling
problem}. Specifically, the model required estimating more ``unknown''
parameters than the number of available pairwise comparisons
\citep[pp.~267]{Thurstone_1927b}. For instance, in a CJ experiment with
five texts, the general form would require estimating \(20\) parameters:
five modal discriminal processes, five discriminal dispersions, and
\(10\) correlations --one per comparison (see
Table~\ref{tbl-thurstone_cases}). However, a single judge could only
provide \({5 \choose 2} = 10\) unique comparisons, an insufficient data
set to estimate the required parameters.

@@

However, Thurstone designed Case V to provide a ``rather coarse
scaling'' of traits \citep[pp.~269]{Thurstone_1927b}, prioritizing
statistical simplicity over precise trait measurement. He cautioned that
its use ``should not be made without (an) experimental test''
\citep[pp.~270]{Thurstone_1927b} because it imposes the most extensive
set of simplifying assumptions among all the proposed cases
\citetext{\citealp[pp.~253]{Bramley_2008}; \citealp[pp.~677]{Kelly_et_al_2022}}
(see Table~\ref{tbl-thurstone_cases}). Therefore, it is surprising that,
despite these limitations, CJ research has predominantly relied on Case
V to measure different traits, raising significant concerns about the
reliability and validity of such measurements in contexts where the
case's assumptions may not hold \citep[pp.~677]{Kelly_et_al_2022}.

Furthermore, since Thurstone's primary goal was to define a
psychological scale and ``allocate the compared stimuli on this
continuum'' \citep[pp.~269]{Thurstone_1927b}, he did not provide
specific guidance on how to use the measurement scores from Case V to
conduct hypothesis tests related to the measured traits. Although the CJ
practice has circumvented this issue by using the point estimates of the
scores or their transformations for such analyses, a fundamental
question remains: Is this method appropriate for addressing these
inquiries?

Thus, this section discusses these two prominent issues. Specifically,
Section~\ref{sec-theory-issue1} examines the heavy reliance on
Thurstone's Case V assumptions in the statistical analysis of CJ data.
Conversely, Section~\ref{sec-theory-issue2} focuses on the apparent
disconnect between the approaches to trait measurement and hypothesis
testing in CJ.

\subsection{The Case V and the statistical analysis of CJ
data}\label{sec-theory-issue1}

Despite its limitations, Case V remains the most widely used model in CJ
literature. This preference primarily stems from the BTL model, which
provides a simplified statistical representation of the case. The BTL
model mirrors the assumptions of Case V, with one notable distinction:
while Case V assumes a Normal distribution for the stimuli' discriminal
processes, the BTL model uses the more mathematically tractable Logistic
distribution \citep[pp.~254]{Andrich_1978, Bramley_2008} (see
Table~\ref{tbl-thurstone_cases}). However, as
\citet[pp.~16]{vanderLinden_et_al_2017_I} noted, this substitution has
minimal impact on the model's estimation or interpretation because the
discriminal process scale is arbitrary up to a non-monotonic
transformation. Furthermore, this limited impact is supported by the
fact that the Normal and Logistic distributions exhibit analogous
statistical properties, differing only by a scaling factor of
approximately \(1.7\).

However, Thurstone acknowledged that some assumptions of Case V could be
problematic when researchers assess complex traits or heterogeneous
stimuli \citep[pp.~376]{Thurstone_1927a}. Thus, given that modern CJ
applications often involve such traits and stimuli, two key assumptions
of Case V, and by extension, the BTL model, may not always hold in
theory or practice. These assumptions are the equal dispersion and zero
correlation between stimuli.

@@

According to the theory, discrepancies in the discriminal dispersions of
stimuli shape the distribution of the discriminal difference, directly
influencing the outcome of pairwise comparisons. A thought experiment
can help illustrate this idea. In this experiment, researchers observe
the discriminal processes for two texts, similar to those shown in
Figure~\ref{fig-discriminal_process}. Additionally, the discriminal
dispersion for Text A remains constant, and the texts are uncorrelated
\((\rho=0)\). Figure~\ref{fig-dispersion} demonstrates that an increase
in the uncertainty associated with the perception of Text B relative to
Text A \((\sigma_{B} - \sigma_{A})\), broadens the distribution of their
discriminal difference. This broadening affects the probability area
where the discriminal difference distinctly favors Text B over Text A,
expressed as \(P(B>A)\), ultimately influencing the comparison outcome.
Additionally, the figure reveals that when the discriminal dispersions
of the texts are equal, as in the BTL model
\((\sigma_{B} - \sigma_{A}=0)\), the discriminal difference is more
narrow than when the dispersions differ. As a result, the discriminal
difference is more likely to favor Text B over Text A, as it is
represented by the shaded gray area.

@@

In experimental practice, however, the thought experiment occurs in
reverse. Researchers first observe the comparison outcome and then use
the BTL model to infer the discriminal difference between stimuli and
their respective discriminal processes \citep[pp.~373]{Thurstone_1927a}.
Consequently, the outcome's ability to reflect \emph{true} differences
between stimuli largely depends on the validity of the model's
assumptions \citep[pp.~150]{Kohler_et_al_2019}, in this case, the
assumption of equal dispersions. For instance, when the assumption
accurately captures the complexity of the data, the BTL model estimates
a discriminal difference distribution that accurately represents the
\emph{true} discriminal difference between the texts. This scenario is
illustrated in Figure~\ref{fig-dispersion}, when the model's discriminal
difference distribution aligns with the \emph{true} discriminal
difference distribution, represented by the thick continuous line
corresponding to \(\sigma_{B}-\sigma_{A}=0\). The accuracy of this
discriminal difference then ensures reliable estimates for the texts'
discriminal processes.

Notably, while the assumption of equal dispersions simplifies the trait
measurement process, evidence from the CJ literature suggests that the
BTL model may fail to capture the complexity of modern CJ data.
Specifically, the assumption of equal dispersions may not hold when
researchers assess complex traits or heterogeneous stimuli
\citetext{\citealp[pp.~376]{Thurstone_1927a}; \citealp[pp.~273]{Bramley_2008}; \citealp[pp.~678]{Kelly_et_al_2022}},
as these traits and stimuli can introduce judgment discrepancies due to
their unique characteristics
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}.
Indeed, the CJ literature may already provide evidence of such
discrepancies, particularly in the form of ``misfit'' statistics. For
instance, \emph{misfit texts} are those whose comparisons result in more
judgment discrepancies than those involving other texts
\citetext{\citealp[pp.~11]{Pollitt_2004}; \citealp[pp.~290]{Pollitt_2012b}}.

These discrepancies may arise from two possible causes. First, the
discriminal processes of misfit texts may substantially overlap with
those of non-misfit texts. Second, misfit texts may represent outlying
observations with characteristics that make them difficult to compare
and position on the trait scale. Regarding the first cause, the overlap
can occur when misfit texts share the same trait space and have equal
dispersion as non-misfit texts or when they exhibit larger discriminal
dispersions. Although the BTL model can distinguish the first scenario,
it is inherently agnostic to the second.
Figure~\ref{fig-discriminal_process} illustrates how an increased
discriminal dispersion can widen the overlap between two texts, leading
to more judgment discrepancies. In the figure, text B has a broader
discriminal dispersion than text A, expanding the overlap between their
discriminal processes, where judges struggle to determine which text
possesses more of the trait.

Thus, erroneously assuming equal dispersions between stimuli can give
rise to significant statistical and measurement issues. Specifically,
the BTL model may overestimate the trait's reliability, that is, the
degree to which the outcome accurately reflects the \emph{true}
discriminal differences between stimuli. This overestimation, in turn,
results in spurious conclusions about these differences
\citep[pp.~370]{McElreath_2020} and, by extension, about the underlying
discriminal processes of stimuli. Figure~\ref{fig-dispersion} also
illustrates this scenario when the model's discriminal difference
distribution aligns with the thick continuous line for
\(\sigma_{B}-\sigma_{A}=0\), while the \emph{true} discriminal
difference follows any discontinuous line where
\(\sigma_{B}-\sigma_{A} \neq 0\).

Finally, regarding the second cause, if researchers recognize that
misfit statistics may represent outlying observations, the conventional
CJ practice of excluding stimuli based on these statistics
\citep{Pollitt_2012a, Pollitt_2012b, vanDaal_et_al_2016, Goossens_et_al_2018},
may unintentionally discard valuable information, introducing bias into
the trait estimates \citep[chap.~12]{Zimmerman_1994, McElreath_2020}.
The direction and magnitude of these biases remain unpredictable, as
they depend on which stimuli researchers exclude from the analysis.

@@

The correlation \(\rho\) measures how much the judges' perception of a
specific trait in one stimulus depends on their perception of the same
trait in another. As with the discriminal dispersions, this correlation
shapes the distribution of the discriminal difference, directly
impacting the outcomes of pairwise comparisons. A similar thought
experiment, as the one depicted in Section~\ref{sec-theory-issue1a}, can
illustrate this idea. The experiment only assumes that the discriminal
dispersions for both texts remain constant. Figure~\ref{fig-correlation}
reveals that as the correlation between the texts increases, the
distribution of their discriminal difference becomes narrower. This
narrowing affects the area under the curve where the discriminal
difference distinctly favors Text B over Text A, denoted as
\(P(B > A)\), thus influencing the comparison outcome. Furthermore, the
figure shows that when two texts are independent or uncorrelated
\((\rho=0)\), their discriminal difference is less narrow compared to
scenarios where the texts are positively correlated. As a result, the
discriminal difference is less likely to favor Text B over Text A, as it
is represented by the shaded gray area.

As outlined in the previous section, researchers approach this process
in reverse. They begin by observing the outcomes and using the BTL model
to estimate the stimuli' discriminal differences and discriminal
processes. Given that the BTL model assumes independent discriminal
processes across comparisons, if this assumption holds in the data, the
model estimates a discriminal difference distribution that accurately
reflects the \emph{true} discriminal difference between the texts. Once
more, the accuracy of the discriminal difference ensures reliable
estimates for the discriminal processes of the texts.

Thurstone assumed that stimuli were uncorrelated because judges' biases,
arising from two opposing and equally weighted effects occurring during
the pairwise comparisons, canceled each other out
\citep[pp.~268]{Thurstone_1927b}. \citet{Andrich_1978} provided a
mathematical demonstration of this cancellation using the BTL model
under the assumption of discriminal processes with additive biases.
However, evidence from the CJ literature indicates that the assumption
of zero correlation does not hold in practice in at least two scenarios:
when intricate aspects of multidimensional, complex traits or
heterogeneous stimuli influence judges' perceptions or when additional
hierarchical structures are relevant to the stimuli.

In the first scenario, research on text quality suggests that when
judges evaluate multidimensional, complex traits or heterogeneous
stimuli, they often rely on various intricate characteristics of the
stimuli to form their judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
additional relevant characteristics, when assessed, are unlikely to be
equally weighted or opposing. As a result, they may exert an uneven
influence on judges' perceptions, creating biases that resist
cancellation. For example, this could occur when a judge assessing the
argumentative quality of a text places more weight on its grammatical
accuracy than other judges, thereby favoring texts with fewer errors but
weaker arguments. Furthermore, since the discriminal difference of the
stimuli becomes an observable outcome only through the judges'
perceptions, these biases could introduce dependencies between the
stimuli \citep[pp.~346]{vanderLinden_et_al_2017_II}. While direct
evidence for this particular scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
supporting the notion that the factors influencing pairwise comparisons
may not always cancel out.

In the second scenario, the shared context or inherent connections
introduced by additional hierarchical structures may create dependencies
between stimuli, a statistical phenomenon known as clustering
\citep{Everitt_et_al_2010}. Although the CJ literature acknowledges the
existence of such hierarchical structures, the statistical approaches to
account for this additional source of dependence have been insufficient.
For instance, when CJ data incorporates multiple samples of stimuli from
the same individuals, researchers frequently rely on (averaged) point
estimates of the BTL scores to conduct subsequent analyses and tests at
the individual level
\citep{Bramley_et_al_2019, Boonen_et_al_2020, Bouwer_et_al_2023, vanDaal_et_al_2017, Jones_et_al_2019, Gijsen_et_al_2021}.
However, this approach can introduce additional statistical and
measurement issues, which we discuss in greater detail in
Section~\ref{sec-theory-issue2}.
